Continuous Deductive Reasoning over Streamed RDF Data

### 2015-07-10 content/contribution/continuous-datalog.tex
Revised definitions for time

A conventional Datalog program $\Pi = \langle R, F\rangle$ is a tuple
which comprises a finite set of rules $R$ and a finite set of ground
facts $F$. Facts are extensional atomic formulae of first-order logic,
and rules are of the form:

\[\alpha_0 :- \alpha_1 , \ldots , \alpha_n\]
where $\alpha_0$ is an intensional atomic formula and
$\alpha_1 \ldots \alpha_n$ are a mixture of intensional and
extensional atomic formulae.

For the purpose of defining Continuous Datalog in the context of the
Semantic Web, we treat facts as 3-tuples $\langle s, p, o \rangle$,
where $s, p, o$ are respectively the subject, predicate and object of
an RDF triple. 
\end{definition}

\begin{definition}[Entailment Justification]
\labeldef{continuous datalog: justification}
A justification for some entailment $f$ of some Datalog program
${\Pi = \langle R , F \rangle \vdash f}$ is a minimal set ${J \subseteq F}$
such that ${\Pi^\prime = \langle R , J \rangle \vdash f}$.
\end{definition}

\begin{definition}[Instant]
\labeldef{continuous datalog: instant}
An instant is a discrete point in time, such that an instant 
${i \in \mathbb{N}}$.

It should be noted that time is a continuous concept, with infinitely
many divisions possible between any two discrete instants.  However,
the resolution with which such divisions can be recognised will, in
practise, have a lower bound set by hardware limitations, or may be
defined as part of some specific implementation.  In these cases,
anything that occurs between discrete instants is considered to have
occurred at the next recognised instant.
\end{definition}

\begin{definition}[Fact Instance]
\labeldef{continuous datalog: fact instance}
A {\em fact instance} is a fact with some additional temporal annotation;
it is the combination of a fact's signature (its predicate $p$, subject
$s$ and object $o$) and the timestamp $t$ at which the instance became true.
A fact instance is denoted ${\langle s, p, o \rangle | t \in \mathbb{F^T}}$.

We represent the relationship between a fact (without temporal
annotation) and a fact instance (with temporal annotation) with the
function ${strip : \mathbb{F^T} \rightarrow \mathbb{F}}$ which maps a fact instance to its
underlying fact ${\langle s, p, o \rangle \in \mathbb{F}}$.
\end{definition}

@@

\begin{definition}[Sliding Window]
\labeldef{continuous datalog: window}
A sliding window of range ${x \in \mathbb{N}}$ over any stream $S$ at
any given instant ${i \in \mathbb{N}}$ is the \emph{set} of fact
instances that occur in $S$ during the interval ${(i-x,i]}$, and is
denoted ${W^{S,x}_{i}}$.
\end{definition}

@@

In \refdef{continuous datalog: datalog program}, we took a
Datalog program to consist of a set of rules and a set of facts.
Taken together with the set of rules $R$, the set of facts $F$ entails
another set of facts $E$ in which the existence of one fact in $E$ may
be justified by many distinct subsets of $F$. In order to extend this
definition of Datalog programs to incorporate fact insatnces,
as in \refdef{continuous datalog: fact instance}, we define a
time-annotated Datalog program in \refdef{continuous datalog: ta datalog program}.

>>>>

A conventional Datalog program $\Pi = \langle R, F\rangle$ comprises a
set of clauses, divided into a finite set of rules $R$ and a finite
set of ground facts $F$. Facts are formulae of first-order logic, and
rules are of the form:

\[\alpha_0 :\!-\ \alpha_1 , \ldots , \alpha_n\]
where $\alpha_i$ are atomic formulae.

The Herbrand base $\hat{H}$ of $\Pi$ is the set of all facts that can
be expressed using the predicates in $\Pi$; for a program $\Pi =
\langle R, F \rangle$, $F \subseteq \hat{H}$. Note that we do not
distinguish between the predicates in $F$ and those in the heads of
the rules in $R$ (as extensional and intensional predicates), as is
the case in the model theory for Datalog presented in
\cite{datalog-basics}; in this respect, our treatment follows that of
the model theory for Prolog presented in \cite{prolog-semantics}. A
Herbrand interpretation $\mathcal{I}$ of $\Pi$ is any subset of
$\hat{H}$, and a Herbrand interpretation $\mathcal{I}$ that satisfies
the set of clauses in program $\Pi$ is a Herbrand model of $\Pi$,
written $\models_{\mathcal{I}} \Pi$.

We define the concept of entailment in Datalog as follows. A fact $f$
is entailed by a program $\Pi$ iff each interpretation satisfying
$\Pi$ (satisfying every clause in $\Pi$) also satisfies $f$, written
$\Pi \models f$. The entailment of $\Pi$ is therefore the set of all
facts that logically follow from $\Pi$:

\[cons(\Pi) = \{ f \in \hat{H} | \Pi \models f \}\]
\end{definition}

\begin{definition}[Justification]
\labeldef{continuous datalog: justification}
A {\em justification} for the entailment $E$ of some Datalog program
${\langle R , F \rangle \models E}$ is a minimal set
${J \subseteq F}$ such that ${\langle R , J \rangle \models E}$.
\end{definition}

\begin{definition}[Instant]
\labeldef{continuous datalog: instant}
We assume a discrete, ordered time domain $\mathbb{T}$; an instant $i
\in \mathbb{T}$ is a point in time. It should be noted that time is a
continuous concept, with infinitely many divisions possible between
any two discrete instants.  However, the resolution with which such
divisions can be recognised will, in practise, have a lower bound set
by hardware limitations, or may be defined as part of some specific
implementation.  In these cases, anything that occurs between discrete
instants is considered to have occurred at the next recognised
instant.
\end{definition}

\begin{definition}[Fact Instance]
\labeldef{continuous datalog: fact instance}
A {\em fact instance} is a fact with some additional temporal
annotation; it is a tuple containing a fact $f$ and the timestamp
$t \in \mathbb{T}$ at which $f$ became true, written
$\langle f, t \rangle$. Two fact instances containing the same fact
and timestamp are considered identical.

We represent the relationship between a fact (without temporal
annotation) and a fact instance (with temporal annotation) with the
function ${strip : \hat{H} \times \mathbb{T} \rightarrow \hat{H}}$
which maps a fact instance to its underlying fact.
\end{definition}

@@

\begin{definition}[Sliding Window]
\labeldef{continuous datalog: window}

A sliding window of range ${x \in \mathbb{N}}$ over any stream $S$ at
any given instant ${i \in \mathbb{N}}$ is the \emph{set} of fact
instances that occur in $S$ during the interval ${(i-x,i]}$, written
${W^{S,x}_{i}}$.

\[ W^{s,x}_i = \{ \langle f, t \rangle | \langle f, t \rangle \in S \land (i-x) < t \leqslant i \}\]

\end{definition}

@@

In \refdef{continuous datalog: datalog program}, we took a
Datalog program to consist of a set of rules and a set of facts.
Taken together with the set of rules $R$, the set of facts $F$ entails
another set of facts $E$ in which the existence of one fact in $E$ may
be justified by many distinct subsets of $F$.

In order to extend this
definition of Datalog programs to incorporate fact insatnces,
as in \refdef{continuous datalog: fact instance}, we define a
time-annotated Datalog program in \refdef{continuous datalog: ta datalog program}.
